---
  title: "Having fun with FUNctions ;) <br/>"
author: "Nick Sard <br/> Fisheries and Wildlife, OSU, HMSC, Newport, OR <br/>"
date: "Presented 1a4 April, 2015, HMSC UseRs meeting"
output:
  ioslides_presentation:
  fig_height: 5
fig_width: 8
highlight: tango
beamer_presentation:
  highlight: tango
---
  
  ## Sources for some of the code and ideas I used:
  - http://www.ats.ucla.edu/stat/r/library/intro_function.htm
- http://adv-r.had.co.nz/Functions.html
- http://www.statmethods.net/management/userfunctions.html
- http://www.stat.berkeley.edu/~statcur/Workshop2/Presentations/functions.pdf
- http://www.r-bloggers.com/how-to-write-and-debug-an-r-function/
  - http://cran.r-project.org/doc/contrib/Lemon-kickstart/kr_rfunc.html
- http://nicercode.github.io/guides/functions/
  
  ## Outline:
  - What is a function?
- Basic examples
- Returning values
- Local and global values
- Debugging
- Location and organization
- Source()


## What is a function?

- Basic structure

```{r}
name.of.function <- function(argument1, argument2) {
  statements
  return(something)
}
```
- Make the name informative
- Arguments can be single values, vectors, matrices, data frames, lists
- Statements are what you want done
- It can be just a single line of code or many
- Try not to make it too long
- Don't forget the return() at the end!

## Returning values to user
- Make a simple function
```{r}
fun1 <- function(x) {
x*2  
3 * x - 1
}
fun1(5)
```
- Only the last expression evaluated is returned
- Note that we didn't save  the calculation to an object
- So it gets automatically returned to the console

## Returning values to user
- Same function except we saved it to an object
```{r}
fun2 <- function(x) {
  y <- 3 * x - 1
}
fun2(5)
```
- Doesn't return anything because the object was a local variable
- Local variables are defined inside {}
- E.g. For loops and functions

## Returning values to user
- Two ways to resolve this "issue"
- 1. Recommend using return()
```{r}
fun2.1 <- function(x) {
a <- x*2
b <- paste(a, "is my favorite number")
y <- 3 * x - 1
return(y)
}
fun2.1(5)
```
- But a and b were cannot be returned...

## Returns only a single object, but we can use a list to return multiple objects

```{r}
fun2.1 <- function(x) {
a <- x*2
b <- paste(a, "is my favorite number")
y <- 3 * x - 1
return(list(a,b,y))
}
fun2.1(5)
```

## Local and global variables
- Can also save it to a global object
```{r}
fun2.2 <- function(x) {
Y <<- 3 * x - 1
}
fun2.2(6)
```
- Use <<-
- Convention is to use CAPITAL names like Y
- Not recommended - Only seen this occasionally used in for loops

## Some functions use other functions
- Goal is to create a function that welcomes you to R
- There are some functions built into R
- http://www.statmethods.net/management/functions.html
```{r}
#Gives your system's information and the R version you are using
myinfo <- Sys.info() 
myinfo[7]

rinfo <- R.Version()
rinfo$nickname
```

## Now for the function...

```{r}
hiyou <- function(){
  #extracting some information
  myinfo <- Sys.info() ; rinfo <- R.Version()
  
  #making three statements
  hi <- paste("Hi",paste0(myinfo[7],"!"))
  welcome <- paste("Welcome to",paste0(rinfo$version.string,"!"))
  iam <- paste("My name is:",paste0(rinfo$nickname,"."),
               "Let's go have some fun!")
  
  #returning a cat statement
  return(cat(paste0(hi,"\n",welcome,"\n",iam)))
}
hiyou()
```

## Creating a stats summary function 

```{r}
mysummary <- function(x, output=1) {
  #calculating stats of interest
  center <- mean(x) 
  spread <- sd(x)
  
  #Using output to determine how the information is
  #displayed to the user
  if (output == 1) {
    return(cat("Mean=", center, "\n", "SD=", spread, "\n"))
  } 
  
  if (output == 2) {
    df <- data.frame(statistic = c("mean","sd"),
                     value = c(center,spread))  
    return(df)
  }
}
```

## Basic stat function example

```{r} 
set.seed(1234)
x <- rnorm(n = 500,mean =  4,sd = 1) 
mysummary(x)
mysummary(x, output=2)
mysummary(output=2,x)
```

## Debugging
-  We can add some if statements to prevent errors
```{r}
mysummary <- function(x, output=1) {
  #calculating stats of interest
  center <- mean(x) 
  spread <- sd(x)
  #Using output to determine how the information is
  #displayed to the user
  if (output == 1) {
    return(cat("Mean=", center, "\n", "SD=", spread, "\n"))
  } 
  if (output == 2) {
    df <- data.frame(statistic = c("mean","sd"),
                     value = c(center,spread))  
    return(df)
  }
}
```

## Debugging
- Use if() to make sure the right type of data is being used. When it isn't -
- stop() - stops function
- warning() - adds a warning
- print() can provide feedback to user
```{r}
mysummary <- function(x, output=1) {
if(is.numeric(x)!=T){
stop("x is not numeric")
}
if(length(x)<30) {
warning("Sample size is <30",call. = F)
} 
print("Calculating mean and standard deviation")
# the rest of the function...
}
```

## mysummary() in action

```{r,echo=F}
mysummary <- function(x, output=1) {

if(is.numeric(x)!=T){
stop("x is not a vector of numeric values")
}

if(length(x)<30) {
warning("Sample size is less than 30",call. = F)
} 

print("Calculating mean and standard deviation")
#calculating stats of interest
center <- mean(x) 
spread <- sd(x)

#Using output to determine how the information is
#displayed to the user
if (output == 1) {
return(cat("Mean=", center, "\n", "SD=", spread, "\n"))
} 

if (output == 2) {
df <- data.frame(statistic = c("mean","sd"),
value = c(center,spread))  
return(df)
}
}
```

```{r}
mysummary(x = rnorm(100,4,1))
#mysummary(x = LETTERS[1:10])
mysummary(x = c(1,2,3))
```

## Functions with apply()
- Goal calculate geometric mean across all columns

```{r,echo=FALSE}
data(mtcars)
```

```{r}
gm_mean = function(x, na.rm=TRUE){
exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))
}
# column geometric means - apply() takes matrices
apply(as.matrix(mtcars[,1:4]), MARGIN = 2, gm_mean)

#compared to mean
apply(as.matrix(mtcars[,1:4]), MARGIN = 2, mean)
```

## Recursive functions
- Fibonacci sequence example
```{r}
fib <- function(n){
#base case
if(n==1){
return(1)
} else if(n==2) {
return(1)
#inductive case
} else {
a = fib(n-1)
b = fib(n-2)
return(a+b)
}
}
c(fib(1),fib(2),fib(3),fib(4),fib(5),fib(6))
```

## Best practices for writing functions
- Keep functions short
- Split up function into easy to understand chunks
- Can call other functions
- Code is cleaner and easy to test and update 
- Put in comments so that user knows:
- What the inputs to the function are
- What the function does
- What the output(s) are
- Check for errors 
- Test function with simple examples
- Use debugging and error messages

## Location and organization
- Throughout R script
- Recommended at the beginning
- Single or multiple function(s)
- Save in another script just for functions
- Single or multiple function(s)

## Use source() to call a collection of functions
- Make an R script containing all your functions
- Call at the beginning of your scripts

```{r,eval=FALSE}
source("my.functions.R")
```

## Don't forget to provide necessary information in the source script

```{r}
### function.name()

#Description: This function does some really cool stuff! Details...

#Inputs:
# df - a data frame with three columns
#   column 1 - population same
#   column 2 - sample ID
#   column 3 - length

#Outputs
# options - Default is 1
#   option 1 - returns mean and standard deviation for length
#   option 2 - creates a histogram of lengths
```
